---
title: "Data Wrangling with R"
author: "Meenakshi Kushwaha"
date: "21st July, 2022"
format: 
  revealjs:
    theme: [my_theme.scss]
    incremental: true

execute:
  echo: true
---

## Prerequsities

```{r}
library(gapminder) # data set
library(tidyverse) # today's package
```

## Basics

-   Pick observations by their values `filter()`.
-   Reorder the rows `arrange()`.
-   Pick variables by their names `select()`.
-   Create new variables with functions of existing variables `mutate()`.
-   Collapse many values down to a single summary `summarise()`
    -   often used with `group_by()`

## Code {auto-animate="true"}

When you click the **Render** button a presentation will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
a<- 1 + 1
```

## Code {auto-animate="true"}

```{r}
a<- 1 + 1
print(a)

```

## Executable Code

```{r}
#| echo: true
#| fig-width: 10
#| fig-height: 4.5
#| output-location: slide
library(ggplot2)
ggplot(mtcars, aes(hp, mpg, color = am)) +
  geom_point() +  geom_smooth(formula = y ~ x, method = "loess")
```

::: footer
Learn more: [Executable Code](https://quarto.org/docs/presentations/revealjs/#executable-code)
:::

## About the dataset {.smaller}

Arrange content into columns of varying widths:

::: columns
::: {.column width="25%"}
#### Gapminder dataset

Dataset of 142 countries, with values for life expectancy, GDP per capita, and population, every five years, from 1952 to 2007
:::

::: {.column width="13%"}
:::

::: {.column width="62%"}
```{r echo=FALSE}
knitr::kable(head(gapminder))
```
:::
:::

::: footer
Learn more: [Multiple Columns](https://quarto.org/docs/presentations/revealjs/#multiple-columns)
:::

## Gapminder dataset

```{r}
head(gapminder, 15)
```

## Fragments

Incremental text display and animation with fragments:

<br/>

::: {.fragment .strike}
Strike
:::

::: {.fragment .highlight-red}
Highlight red
:::

::: footer
Learn more: [Fragments](https://quarto.org/docs/presentations/revealjs/advanced.html#fragments)
:::

## Speaker notes

::: notes
Speaker notes go here.
:::

## Basic syntax

function(argument 1, argument 2, argument 3, ...) - Result is a new dataframe - functions are nothing but verbs

filter(dataframe, age == "20")

## Filter rows with `filter()`

```{r}
#| echo: true
#| output-location: fragment
filter(gapminder, country == "India")
```
